Topology and experimental distinguishability

### 2017-07-21 Articles/TopologyAndExperimentalDistinguishability/article.tex
possibly finished with core content revisions

Now that we have a framework for testing a system, we will define a class of systems we can understand with these tests. The framework will be general enough that one can readily apply it to many real-world experimental scenarios. In this construction, we will begin with a space we wish to understand - in this case, finding the appropriate element - and then define the appropriate collection of experimental observations to do so. In what follows, let $X$ be a set.

@@

\begin{thm}
An experimentally distinguishable identification problem $(X,\mathcal{D})$ forms a Hausdorff, second-countable topology $(X,\mathsf{T})$ with open sets given by the verifiable sets. 
\end{thm}
\begin{proof}
First, from the definition one can see that for all $x\in X$, there exists a verifiable set $U$ with $x\in U$, so their union $X$ is also a verifiable set. Further, there is a null observation by taking the conjunction of two identifications corresponding to disjoint sets, so the empty set is a verifiable set. Now, Proposition \ref{setbehavior} shows that the collection $\mathsf{T}$ is closed under finite intersection and countable union. Because $\mathsf{T}$ is determined by an experimental domain of observations, there is a countable basis of operations which translates to a countable basis of open (verifiable) sets, so it is second-countable. To show it is closed under arbitrary union, notice that an arbitrary union may be rewritten as a union of basis elements, which is then a countable union, and so it remains in $\mathsf{T}$. That the topology is Hausdorff is immediate from the last part of the definition. 
\end{proof}

And so we have found that all experimentally distinguishable spaces may be given by Hausdorff, second-countable topological spaces.

@@

TODO

Now that we have defined the possibilities of our domains and the topology associated with them, we want to establish relationships between them. The aim of scientific investigation is studying and characterizing these relationships.

\begin{prop}
	\label{setfunctions}
	Let $X$ and $Y$ be two Hausdorff topological spaces, and let $g: \mathsf{T}(Y) \rightarrow \mathsf{T}(X)$ be a mapping such that:
	\begin{enumerate}
		\item It is compatible with union and intersection $\forall V_1, V_2 \in Y$ $g(V_1 \cup V_2)=g(V_1)\cup g(V_2)$ and $g(V_1 \cap V_2)=g(V_1)\cap g(V_2)$
		\item $g(\emptyset) = \emptyset$
		\item $g(Y) = X$
	\end{enumerate}
	Then there exists a unique continuous function $f: X \rightarrow Y$ such that $g = f^{-1} |_{\mathsf{T}(Y)}$.
\end{prop}

\begin{proof}
	We claim there exists a unique extension $\bar{g}:\sigma(Y)\to\sigma(X)$ to the Borel $\sigma$-algebras of $X$ and $Y$, respectively $\sigma(X)$ and $\sigma(Y)$, such that $\bar{g}|_{\mathsf{T}(Y)}=g$ and $\bar{g}$ is compatible with union, intersection and complements. Let $\bar{g}(V) = g(V)$ for all open sets $V \in \mathsf{T}(Y)$. Let $A \in \sigma(Y)$ (not necessarily open) and $A^C$ be its complement. We must have $\bar{g}(A^C) = \bar{g}(A)^C = X\setminus \bar{g}(A)$ for $\bar{g}$ to be compatible with complements. Recall that all Borel sets in $\sigma(Y)$ and $\sigma(X)$ may be written as some combination of unions, intersections, and complements of open sets. Thus, the construction uniquely determine what $\bar{g}$ should output on any Borel set. We need only check that the output is still a Borel set. But by definition of $\bar{g}$, the outputs will be given as unions, intersections, and complements of outputs of $g$, which are open sets, and so the image of $\bar{g}$ is contained in $\sigma(X)$.  $\bar{g}$ is well defined. The function $\bar{g}$ in a sense represents extracting the maximum amount of information possible out of $g$.

@@

For continuous functions to be physically distinguishable, we need to show they can always be given, as a set, given a topology that is Hausdorff and second-countable.

To that end, we introduce the basis-to-basis topology on the set of continuous functions from two topological spaces. This is the topology generated by the sets of functions that map a basis element of one element inside an element of the other space.

>>>>

Now that we have a framework for testing a system, we will define a class of systems we can understand with these tests. The framework will be general enough that one can readily apply it to many real-world experimental scenarios. In this construction, we will begin with a space we wish to understand - in this case, finding the appropriate element - and then define the appropriate collection of experimental observations to do so. The idea is that we have some object, and $X$ is a collection of possible attributes. We wish to identify the appropriate attribute among the collection. For example, the object might be an electron, and the collection of possible attributes might be the set of real numbers, representing possible masses. In what follows, let $X$ be a set.

@@

\begin{thm}
An experimentally distinguishable identification problem $(X,\mathcal{D})$ has a natural Hausdorff, second-countable topology $(X,\mathsf{T})$ with open sets given by the verifiable sets. 
\end{thm}
\begin{proof}
First, from the definition one can see that for all $x\in X$, there exists a verifiable set $U$ with $x\in U$, so their union $X$ is also a verifiable set. Further, there is a null observation by taking the conjunction of two identifications corresponding to disjoint sets, so the empty set is a verifiable set. Now, Proposition \ref{setbehavior} shows that the collection $\mathsf{T}$ is closed under finite intersection and countable union. Because $\mathsf{T}$ is determined by an experimental domain of observations, there is a countable basis of operations which translates to a countable basis of open (verifiable) sets, so it is second-countable. To show it is closed under arbitrary union, notice that an arbitrary union may be rewritten as a union of basis elements, which is then a countable union, and so it remains in $\mathsf{T}$. That the topology is Hausdorff is immediate from the last part of the definition. 
\end{proof}

We refer to the topological space $(X,\mathsf{T})$ resulting from an experimentally distinguishable identification problem as an \textbf{experimentally distinguishable space}. And so we have found that all experimentally distinguishable spaces may be given by Hausdorff, second-countable topological spaces.

@@

Now that we have defined the possibilities of our domains and the topology associated with them, we want to establish relationships between experimentally distinguish. The aim of scientific investigation is studying and characterizing these relationships, and so we provide a formalism which not only gives a clear definition of experimental relationships, but we find that experimental relationships themselves can be given the structure of an experimentally distinguishable space.

First, suppose we are given a function $f:X\to Y$, where $Y$ is an experimentally distinguishable space $(Y,\mathsf{T}_Y)$ whose distinguished element is $y^*$, and suppose there exists $x^*\in X$ with $f(x)=y^*$.

\begin{prop}
For each $\mathsf{o}_Y = \llparenthesis y^*\in U_Y,\mathsf{e}_{\in}(U_Y)\rrparenthesis$, there exists an induced experimental identification $\mathsf{o}_X = \llparenthesis x^*\in f^{-1}(U_Y), \mathsf{e}_{\in}(f^{-1}(U_Y)) \rrparenthesis$.
\end{prop}
\begin{proof}
We wish to show that $\mathsf{e}_{\in}(f^{-1}(U_Y))$ is an experimental test which succeeds if and only if $x^*\in f^{-1}(U_Y)$. To run this test, instead we run $\mathsf{e}_{\in}(U_Y)$ on $y^*$. We show first that if this test succeeds, then the statement $x^*\in f^{-1}(U_Y)$ is true. This test succeeds if and only if $y^*\in U_Y$. But $y^* \in U_Y$ if and only if $f^{-1}(\{y^*\})\subseteq f^{-1}(U_Y)$. Since $f(x^*)=y^*$, this implies $x^*\in f^{-1}(U_Y)$.

Next, if $x^*\in f^{-1}(U_Y)$, then we have $y^*=f(x^*) \in U_Y$, and so the test $\mathsf{e}_{\in}(U_Y)$ succeeds. Hence, we have the observation $\llparenthesis x^*\in f^{-1}(U_Y), \mathsf{e}_{\in}(f^{-1}(U_Y))\rrparenthesis$.

\end{proof}

We call the induced observation the \textbf{pullback} $\mathsf{o}_X = f^*\mathsf{o}_Y$ of the observation $\mathsf{o}_Y$ via $f$. The following corollary is an immediate consequence, and gives the most useful interpretation of continuity in the context of experimentally distinguishable spaces.

\begin{cor}
Let $(X,\mathsf{T}_X)$ and $(Y,\mathsf{T}_Y)$ be experimentally distinguishable spaces with respective experimental domains $\mathcal{D}_X$ and $mathcal{D}_Y$, and let $f:X\to Y$ be a map with $f(x^*)=y^*$. Then $f$ is continuous if and only if all pullbacks of observations from $\mathcal{D}_Y$ are observations in $\mathcal{D}_X$.
\end{cor}

Next, we consider the opposite problem. Suppose we have two experimentally distinguishable spaces $(X,\mathsf{T}_X)$ and $(Y,\mathsf{T}_Y)$, and that we have a well-behaved (with respect to conjunction and disjunction) map on the experimental domains. Then this translates to a map of topologies, taking the verifiable (open) sets of $Y$ to those of $X$, which respects (arbitrary) union and (finite) intersection. In this case, we will show that one can find a continuous map which induces this map of topologies.

\begin{prop}
	\label{setfunctions}
	Let $X$ and $Y$ be two Hausdorff topological spaces, and let $g: \mathsf{T}_Y \rightarrow \mathsf{T}_X$ be a mapping such that:
	\begin{enumerate}
		\item It is compatible with union and intersection, i.e. for any subsets $V_1, V_2 \subseteq Y$, we have $g(V_1 \cup V_2)=g(V_1)\cup g(V_2)$ and $g(V_1 \cap V_2)=g(V_1)\cap g(V_2)$
		\item $g(\emptyset) = \emptyset$
		\item $g(Y) = X$
	\end{enumerate}
	Then there exists a unique continuous function $f: X \rightarrow Y$ such that $g = f^{-1} |_{\mathsf{T}_Y}$.
\end{prop}

\begin{proof}
	We claim there exists a unique extension $\bar{g}:\sigma(Y)\to\sigma(X)$ to the Borel $\sigma$-algebras of $X$ and $Y$, respectively $\sigma(X)$ and $\sigma(Y)$, such that $\bar{g}|_{\mathsf{T}_Y}=g$ and $\bar{g}$ is compatible with union, intersection and complements. Let $\bar{g}(V) = g(V)$ for all open sets $V \in \mathsf{T}_Y$. Let $A \in \sigma(Y)$ (not necessarily open) and $A^C$ be its complement. We must have $\bar{g}(A^C) = \bar{g}(A)^C = X\setminus \bar{g}(A)$ for $\bar{g}$ to be compatible with complements. Recall that all Borel sets in $\sigma(Y)$ and $\sigma(X)$ may be written as some combination of unions, intersections, and complements of open sets. Thus, the construction uniquely determine what $\bar{g}$ should output on any Borel set. We need only check that the output is still a Borel set. But by definition of $\bar{g}$, the outputs will be given as unions, intersections, and complements of outputs of $g$, which are open sets, and so the image of $\bar{g}$ is contained in $\sigma(X)$.  $\bar{g}$ is well defined. The function $\bar{g}$ in a sense represents extracting the maximum amount of information possible out of $g$.

@@

For the collection of continuous functions to be experimentally distinguishable, we need a Hausdorff, second-countable topology. To that end, we introduce the basis-to-basis topology on the set of continuous functions from two topological spaces. This is the topology generated by the sets of functions that map a basis element of one element inside an element of the other space.
